---
title: "untargeted-wf"
output: html_document
---

# Untargeted Workflow for Untargeted Experiments.
Author: Jeremy Hack
```{r}
# Load libraries
library(xcms)
library(RANN)
library(modeest)
library(CAMERA)
library(tools)
library(parallel)
library(pander)
library(magrittr)
library(scales)
library(plyr)
library(dplyr)
library(ggpubr)
library(sm)
library(gtools)
library(plotly)
library(htmltools)
library(tidyr)
#library(MSeasy) Removed from CRAN
library(berryFunctions)
library(fda)
```


Set working directory in `pathToProject`:
```{r message=FALSE}
pathToProject <- "/Users/don/OneDrive\ -\ The\ University\ of\ Melbourne/MStractor-demo"
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir=pathToProject)
```

## Input Params
```{r}
# Specify a "typical" MS file which can be considered a reference. Leave blank for MStractorDemo.
ref <- "./MSfiles/Pooled/pooled_R1.mzXML"
#QCdataAll <- TRUE # IF TRUE, EICs for every molecular feature will be plotted
CPUs <- "max"      # Concurrent CPU threads to run. "max" to autodetect
updatePackages <- "FALSE" # If TRUE, all packages will be automatically updated
fastmode <- "FALSE" # If FALSE, does not generate any .pngs

## Chromatography parameters, tuned to faahKO data (Agilent 1100 400 bar HPLC)
rtStart <- 1    # Start region of interest (in seconds)
rtEnd <- "max"  # End region of interest (in seconds). "max" for RT full range
FWHM_min <- 10  # FWHM in seconds of narrowest peak (10)
FWHM_max <- 90  # FWHM in seconds of broadest peak (40)
rtDelta <- 3705-3673   # Max observed difference in retention time (s) across all
# samples (peak apex to peak apex delta).


## MS parameters, tuned for faahKO data (Agilent MSD SL ion trap)
mzPol <- "negative" # Set to "positive" or "negative" ion mode
mzStart <- 100  # Start of m/z region of interest
mzEnd <- 1650   # End of m/z region of interest
mzErrAbs <- 0.01 # Max m/z delta expected for the same feature across all samples
mzZmax <- 3     # Max charge state expected
EICsMax <- 30   # Max number of chrom. peaks expected for a single EIC
sens <- 1      # Factor (between 0 and 1) for peak extraction sensitivity
# Impacts peak picking thresholds, RAM & CPU utilisation.
# Start with ~0.5.
fileType <- "mzXML" # MS data file type e.g. "mzData", "mzML", "mzXML", "CDF"
```


## Create Dataset
```{r}
runtimeStart<- Sys.time()
source("untargeted-wf-lib.R")

if (CPUs == "max"){
  CPUs <- detectCores(all.tests= TRUE, logical= TRUE)
}
print(paste0(CPUs, " CPUs detected."))

# WARNING: this can be shortened? A lot of repeated code.
if (basename(pathToProject) == "MStractorDemo"){
  datapath <- file.path(find.package("faahKO"), "cdf")
  rawfiles <- dir(datapath, full.names=TRUE, pattern="\\.CDF", recursive=TRUE)
  ref <- paste(datapath,"/WT/wt19.CDF", sep= "")
} else if (fileType == "CDF"){
  datapath <- paste(pathToProject,"MSfiles" , sep="/")
  rawfiles <- dir(datapath, full.names=TRUE,
                  pattern="\\.CDF", recursive=TRUE)
} else if (fileType == "mzData"){
  datapath <- paste(pathToProject,"MSfiles" , sep="/")
  rawfiles <- dir(datapath, full.names=TRUE,
                  pattern="\\.mzData", recursive=TRUE)
} else if (fileType == "mzML"){
  datapath <- paste(pathToProject,"MSfiles" , sep="/")
  rawfiles <- dir(datapath, full.names=TRUE,
                  pattern="\\.mzML", recursive=TRUE)
} else if (fileType == "mzXML"){
  datapath <- paste(pathToProject,"MSfiles" , sep="/")
  rawfiles <- dir(datapath, full.names=TRUE,
                  pattern="\\.mzXML", recursive=TRUE)
} else (cat("No raw files found!\n\nSupported filetypes include:
            \n CDF \n mzData \n mzML \n mzXML", fill= FALSE)
)

cat("The following MS files have been loaded:\n", rawfiles, fill= TRUE, sep="")

```
### Define file name

```{r}
files<-basename(list.files(path="./MSfiles", recursive = TRUE, full.names=FALSE))
Files<-gsub(".mzXML", "", files)
```

Load the reference file specified in `ref` & define the scan range. Set mz step size for seeking new EIC traces.
```{r}
profmethod <- "bin"
profStep <- mzErrAbs*4
refRaw <- xcmsRaw(ref, profstep= profStep, includeMSn= FALSE, mslevel= NULL,scanrange= NULL)
```

Plot TIC inline, and save it as a `.png` to a `/QC` folder. If the `/QC` folder doesn't exist, one will be created. 
```{r}
if (fastmode == TRUE) {
  dir.create("./QC")
  graphics.off()
  pdf("./QC/Ref_TIC.pdf")
  plotTIC(refRaw, ident= FALSE, msident= FALSE)
  dev.off()
}
```

*Edit*: `scanStart` and `scanEnd` should be user-input params, instead of being automatically read off `refRaw@scantime`. `scanStart` and `scanEnd` are global params that don't change after you've read them off the TIC once at the start.

Qn for Luca: "Quantile-driven threshold selection" for `intThresh`?
```{r}
# Determine scan range (indices)
# scanStart := index of first scantime > rtStart and < rtEnd
scanStart <- head(which(refRaw@scantime > rtStart & refRaw@scantime < rtEnd), n=1)

if (identical(rtEnd, "max")) {
  scanEnd <- max(refRaw@scanindex)
  rtEnd <- refRaw@scantime[which(refRaw@scanindex == scanEnd)]
} else {
  scanEnd <- tail(which(refRaw@scantime > rtStart & refRaw@scantime < rtEnd), n=1)
}

print(paste0("scanStart=", scanStart, "; scanEnd=", scanEnd))

# Calculate peak picking parameters
pwMin <- FWHM_min*1.3
pwMax <- pwMin*1.3
mzErrPpmMin <- mzErrAbs/2/mzEnd*1000000
mzErrPpmMax <- mzErrAbs/2/mzStart*1000000
mzErrPpmMean <- mean(c(mzErrPpmMin,mzErrPpmMax))
mzdiff <- mzErrAbs/5
intThresh <- as.integer(quantile(refRaw@env$intensity,1-sens)*10) 
snThresh <- 30/sens
integ <- 1
fitGauss <- TRUE
sleep <- 0
```

Find peaks. 
```{r}
refPks <- findPeaks(refRaw, 
                    method='centWave', 
                    ppm=mzErrPpmMin*2,
                    peakwidth=c(pwMin, pwMax), 
                    snthresh=snThresh,
                    prefilter=c(5,intThresh), 
                    mzCenterFun="mean",
                    integrate=integ, 
                    mzdiff=mzdiff, 
                    verbose.columns= TRUE,
                    fitgauss=fitGauss, 
                    noise=intThresh, 
                    sleep=sleep)
```
Plot picked peaks, and save.
```{r}

#dev.off()
#Pks <- refPks[,c("rt","mz","maxo","into","intb","sn","egauss")]
#write.table(Pks, file= "./Pks.tsv", sep="\t")
if (fastmode==TRUE) {
  pdf("./QC/Ref_EICs_100.pdf")
  #png("./QC/Ref_EICs_100.png", width = 1024, height = 768, units = "px")
  plotPeaks(refRaw, refPks,  c(10,10), width = FWHM_min*10)
  dev.off()
}
```
With the new script first an `XCMSnEXP` object is created , then it is necessary to translate that in an XCMS object and check the sampclass using Centwave. N.B. Whilst Centwave does not perform binning (it uses the raw data directly), downstream methods such as `getEIC` will use the step parameter.

```{r}
cladir<-dirs('./MSfiles//', full.names=FALSE)

dlf<-vector()
for (i in 1:length(cladir)){
  detfls<-length(list.files(path =(paste(pathToProject,'/MSfiles/',cladir[i], sep=''))))
  dlf<-c(dlf,detfls)
}

samplegroup<-vector()
for (i in 1:length(dlf)){
  sg<-rep(cladir[i],dlf[i])
  samplegroup<-c(samplegroup,sg)
}

cols <- rainbow(length(cladir))
ClassCol<-vector()
for (i in 1:length(dlf)){
  sc<-rep(cols[i],dlf[i])
  ClassCol<-c(ClassCol,sc)
}

pd <- data.frame(sample_name = sub(basename(rawfiles), pattern = ".mzXML",
                                   replacement = "", fixed = TRUE),
                 sample_group = samplegroup,
                 stringsAsFactors = FALSE)

raw_data <- readMSData(files = rawfiles, pdata = new("NAnnotatedDataFrame", pd), mode = "onDisk")
```

Generate BPC and TICs from all files (`raw_data`). 
```{r message=FALSE}
png("./QC/overlaid_BPC.png", width = 1024, height = 768, units = "px")
bpis <- chromatogram(raw_data, aggregationFun = "max")
plot(bpis, col = ClassCol)
legend("topright", box.lwd = 2, legend=c(pd$sample_group), pch=c(15, 15,15, 16, 16,16,17,17,17),
col =ClassCol, xpd=FALSE)
dev.off()

png("./QC/overlaid_TIC.png", width = 1024, height = 768, units = "px")
bpis <- chromatogram(raw_data, aggregationFun = "sum")
plot(bpis, col = ClassCol)
legend("topright", box.lwd = 2, legend=c(pd$sample_group), pch=c(15, 15,15, 16, 16,16,17,17,17),
col =ClassCol, xpd=FALSE)

dev.off()
```
Peak detection and creation of an `XCMSnEXP` object.
```{r}
cwp <- CentWaveParam(peakwidth = c(30, 90), 
                     snthresh = snThresh, 
                     prefilter= c(3,intThresh),  
                     integrate = integ, 
                     fitgauss = FALSE, 
                     noise = 0, 
                     verboseColumns = FALSE, 
                     firstBaselineCheck = TRUE, 
                     ppm=6.06, 
                     mzCenterFun= "wMean")### peak det param are the same used in the old script 
xdata <- findChromPeaks(raw_data, param = cwp)
spn<-xdata@phenoData$sample_name
xset <- as(xdata, "xcmsSet")# turns the XCMSnExp object into a xcms set object

sampclass(xset)<-factor((xdata$sample_group))
sampnames(xset)<-spn


xset # To checkout slots use slotNames(xset); ls(xset@groups)
```

Determine the lowest level of replication within a class?
```{r}
classes <- list.dirs(datapath, full.names= FALSE, recursive= FALSE)
classSize <- vector(length= length(classes))
dirs <- list.dirs(datapath, full.names= TRUE, recursive= FALSE)
lst <- vector("list", length(dirs))
names(lst) <- dirs
for (i in 1:length(dirs)){
  classSize[i] <- length(dir(dirs[i]))
}
minClassSize <- min(classSize)
#medClassSize <- median(classSize)
#aveClassSize <- ave(classSize)
#modeClassSize <- mfv(classSize)

#bw <- FWHM_min/2
mzWid <- mzErrAbs
bw <- mzWid/2
minsamp <- ceiling(minClassSize)-1


# group-methods {xcms}
# A peak group is considered valid when it contains either, the minimum number
# of samples set by minSamp, OR at least one sample class satisfies minFrac.

# See groupChromPeaks(), which uses the PeakDensityParam() object as param input, 
# and adjustRtime(), which uses PeakGroupsParam()

xset <- group(
   xset, method= "density", bw= 3, minfrac= 0.3, minsamp= minsamp,
   mzwid= mzWid, max= EICsMax, sleep= 0)

```

```{r}
getTICs(xcmsSet= xset, pngName= "./QC/TICs_Raw.png", rt= "raw")

# Get EICs
xset_grps <- xset@groups
# optional  write.table(xset@groups,file= "./QC/xsgroup.tsv", sep= "\t")
eicRange <- 2*mean(c(FWHM_min,FWHM_max))
eicsRaw <- getEIC(xset, mzrange=xset_grps, rtrange= eicRange ,
                  groupidx = 1:nrow(xset_grps), rt= "raw")

# Plot as individual PNGs
if (fastmode == TRUE) {
  dir.create("./QC/EICs_Raw/")
  do.call(file.remove,list(list.files("./QC/EICs_Raw", full.names= TRUE)))
  graphics.off()
  png(file.path("./QC/EICs_Raw/%003d.png"), h=768, w=1024)
  plot(eicsRaw, xset)
  dev.off()
}
```


```{r}
# Export matrix prior to RT Alignment -------------------------------------
xsTable <- peakTable(xset, filebase= "./QC/xsTable", method= "medret",
                     value= "maxo")

write.table(xsTable[with(xsTable, order(rt, mz)), ],
            file= "./QC/xsTable.tsv", sep= "\t", quote= FALSE, col.names= NA)
```

# RT Alignment
```{r}
align_ref <- match(basename(ref),basename(rawfiles[]))

pdf("./QC/rtAlignLoess.pdf", w=1280, h=1024)
xsAlign <- retcor(xset, method= "loess", missing= 3, extra= 0, span= 0.3,
                  family= "gaussian", plottype= "deviation")
dev.off()
xsAlign
```

```{r}
xsAlign <- group(
  xsAlign, method= "nearest", mzVsRTbalance= 10, mzCheck= mzErrAbs,
  rtCheck= rtDelta, kNN=10)
xsAlign
```

```{r}
# Plot TICs
getTICs(xcmsSet= xsAlign, pngName= "./QC/TICs_Aligned.png", rt= "corrected")
```

```{r}
xsFilled <- fillPeaks(xsAlign, method="chrom", nSlaves=CPUs)

# Plot EICs
xsFilledGrps <- xsFilled@groups
eicsFilled <- getEIC(xsFilled, mzrange= xsFilledGrps, rtrange= FWHM_max*2,
                     groupidx= 1:nrow(xsFilledGrps), rt= "corrected")


dir.create("./QC/EICs_Aligned/")
do.call(file.remove,list(list.files("./QC/EICs_Aligned", full.names= TRUE)))
if (fastmode == TRUE) {
  graphics.off()
  png(file.path("./QC/EICs_Aligned/%003d.png"), h=768, w=1024)
  
  plot(eicsFilled, xsFilled)
  dev.off()
}
```

Reconstruct Spectra - Construct deconvoluted spectra
```{r}
ppm <- mzErrPpmMean/2
mzabs <- mzErrAbs
minfrac <- minClassSize/length(rawfiles)
xs <- xsFilled # Choose an xcmxSet object

# Using CAMERA:
xs_an <- xsAnnotate(xs, polarity= mzPol, nSlaves= 1) # nSlaves>1 is broken
xs_an <- groupFWHM(xs_an, sigma= 6, perfwhm= 1, intval= "maxo")

xs_an <- findIsotopes(xs_an, maxcharge=mzZmax, maxiso=4, ppm= ppm,
                      mzabs= mzabs, intval="maxo", minfrac=minfrac,
                      filter= TRUE)
xs_an <- groupCorr(xs_an, cor_eic_th= 0.7, pval=0.1,
                   graphMethod="hcs", calcIso= TRUE, calcCiS = TRUE,
                   calcCaS= TRUE, cor_exp_th= 0.7)

PksAn <- getPeaklist(xs_an, intval="maxo")

#PksAn <- PksAn[order(as.numeric(PksAn$rt)),]

write.table(PksAn, file=paste("./QC/Pks_An", "tsv", sep="."), sep= "\t",
            col.names= NA, row.names= TRUE)
```

Export Matrix - Filter peaks and export basepeak matrix
```{r}
# 1. Filter lone groups.
# Count number of entities for each factor
pcgroups <- table(PksAn$pcgroup)
min_ions <- 2
PksAnFilt <- droplevels(PksAn[PksAn$pcgroup %in% names(pcgroups)
                              [pcgroups >= min_ions],,drop=FALSE])

# 2. Optionally, select signals that are part of an isotope group
#PksAnFilt <- PksAnFilt[PksAnFilt$isotopes != "", ]
#PksAnFilt <- PksAn[PksAn$isotopes != "", ]

# 3. Determine median signal intensity
sNames <- sampnames(xs)
RespMed <- apply(PksAnFilt[c(sNames)], 1, median, na.rm= TRUE)

RespMed <- apply(PksAnFilt[c(11:19)], 1, median, na.rm= TRUE)
PksAnFilt <-cbind(PksAnFilt, RespMed)

# 4. Keep only the most intense signal from each pcgroup
# Other values may be preferred
#RespMed <- PksAnFilt$RespMed
# May be required if factors aren't recognised
# PeakList$pcgroup <- as.factor(PeakList$pcgroup)
BasePks <- PksAnFilt[ RespMed == ave(RespMed, PksAnFilt$pcgroup,
                                     FUN= function(RespMed) max(RespMed))
                      , ]
# BasePks[order(as.numeric(BasePks$pcgroup)),]

write.table(BasePks[with(BasePks, order(rt, mz)), ],
            file=paste("Pks_BPs", "tsv", sep="."), sep= "\t",
            col.names= NA, row.names= TRUE)
```

Fetch basePeak EICs
```{r}
BP_EICs <- paste(sprintf("%03d",as.numeric(rownames(BasePks))),"png", sep= ".")

dir.create("./EICs_BasePeaks/")
do.call(file.remove,list(list.files("./EICs_BasePeaks/", full.names= TRUE)))
for (i in 1:length(BP_EICs)){
  sourceF <- paste("./QC/EICs_Aligned/", BP_EICs[i], sep= "")
  destF <- paste("./EICs_BasePeaks/", BP_EICs[i], sep= "")
  file.copy(from= sourceF, to= destF, overwrite= FALSE)
}

dir.create("./EICs_BasePeaks_Curated/")
do.call(file.remove,list(list.files("./EICs_BasePeaks_Curated/",
                                    full.names= TRUE)))
for (i in 1:length(BP_EICs)){
  sourceF <- paste("./QC/EICs_Aligned/", BP_EICs[i], sep= "")
  destF <- paste("./EICs_BasePeaks_Curated/", BP_EICs[i], sep= "")
  file.copy(from= sourceF, to= destF, overwrite= TRUE)
}

```

# Stop watch
```{r}
sessionInfo <- sessionInfo()

now <- Sys.time()
time <- unclass(as.POSIXlt(now))
timeStamp <-paste(time$year,time$mon, time$mday, time$hour, time$min, sep="_")
imageName <- paste(timeStamp, "RData", sep= ".")
save.image(imageName)
historyName <- paste(timeStamp,"RHistory", sep= ".")
savehistory(historyName)

runtimeEnd <- Sys.time()
runtime <- runtimeEnd - runtimeStart
runtime

gc()
```

